\chapter*{Acknowledgments}
The pursuit of knowledge is a noble aspiration that led humanity through its most epic quests.  

First and foremost, I want to express my gratitude to Pr. Yves Le Traon for allowing me to do my Ph.D. studies at the University of Luxembourg and to help him teach the Big Data course during four semesters.
Yves was a supportive mentor and a wise counselor, even during the most challenging time of my studies.

I also want to say my sincere thanks to Dr.\ Jacques Klein and Dr.\ Tegawende Bissyande for their excellent advice and their collaboration on the field of Android security.
I have learned a lot thanks to their experience and their insights on the world of academia.

My research experiments on Android malware would not have been possible without the work accomplished by the SerVal group on the Androzoo project.
Thank you to SerVal, especially Dr.\ Kevin Allix, for maintaining this malware dataset, releasing it to the research community, and for our collaboration on extending the platform.

Besides, I thank the administrative team of the University of Luxembourg, and Solène Vincens in particular, for the organization of my thesis defense and their assistance during my Ph.D. studies.
Thanks also to the team at the University of Luxembourg responsible for the High-Performance Computing (HPC) platform for letting me run my batch processing scripts.

On the 21st of March 2018, I had the opportunity to present my work to the team at Google responsible for the security of the Android platform.
I am grateful to Pr. Yves Le Traon and Dr.\ Damien Octeau for giving me the chance of visiting Google headquarters in the beautiful city of San Francisco. 

My Ph.D. studies were also a moment to meet and discuss with passionate colleagues and friends.
Thank you to Dr.\ Matthieu Jimenez, Antonin Carette, Sankalp Ghatpande, Dr.\ Kevin Allix, Dr.\ Alexandre Bartel, Dr.\ Nicolas Sannier, Dr.\ François Fouquet, Dr.\ Assaad Moawad, Dr.\ Thomas Hartmann and Dr.\ Grégory Nain for the numerous discussions on the ultimate question of life, the universe, and the best programming language.

Furthermore, I want to thank the creators and contributors of the open source software that I used during my Ph.D. studies: Linux, Ubuntu, GNOME, Vim, Python, Clojure, Datomic, ElasticSearch, PostgreSQL, CouchDB, Jupyter, Celery, \LaTeX{}, and all NumFOCUS projects. 

Finally, I want to dedicate this dissertation to the loves of my life, my dear wife Alexandra Jacob, and our two daughters, Eglantine and Augustine Hurier Jacob.
I also thank my sister, my parents, my great parents, and my parents in law for their interest in my research.
